Beyond the Iceberg Hypothesis: \\Opening the Black Box of Transport Costs

### 2016-10-19 redaction/Presentations/SeminaireIrvine/slides_Irvine.tex
Début Irvine again

\begin{frame}
\frametitle{Motivation}
\begin{itemize}
\item Trade costs: A central role in international economic analysis \vspace{0.1cm}
\begin{itemize}
\item[-] Declining over the second half of the 20$^{th}$ century (Jacks et al., 2008, Novy, 2013) \vspace{0.1cm}
\item[-] But still significant: Average international trade costs = a 74\% markup over production costs (Anderson \& Van Wincoop, 2004)
\end{itemize}
\item What exactly are ``trade costs''?  \vspace{0.1cm}
\begin{itemize}
\item[-] Transaction costs, policy costs, time costs, and transport costs \textit{per se}
\end{itemize}
\item Transport costs: A sizeable share of international trade costs \vspace{0.1cm}
\begin{itemize}
\item[-]  Amount to $\simeq 30\%$ of international trade costs  $\Leftrightarrow$ A 21\% markup over production costs  \vspace{0.1cm}
%\item[-] Elasticity of trade wrt freight costs = -3 (Behar \& Venables, 2011) \vspace{0.1cm}
\end{itemize}
\item[$\Rightarrow$] If much trade policy barriers have been removed, the transport cost component of trade costs remains sizeable \vspace{0.1cm}
\end{itemize}
\textbf{The paper: On international transport costs}

@@

\begin{frame}
\frametitle{Motivation (cont')}
\begin{itemize}
\item Standard modeling of trade costs: As an ad-valorem tax-equivalent \vspace{0.1cm}
\begin{itemize}
\item[-] As a constant percentage of the producer price per unit traded \vspace{0.1cm}
\item[$\Leftrightarrow$] The ``iceberg cost'' hypothesis (Samuelson, 1954) \vspace{0.1cm}
\begin{itemize}
\item[-] With $p$ the import price, $\widetilde{p}$ the export price, $q$ the quantity traded and $\tau >1$ the trade cost \vspace{0.1cm}
\footnotesize
$$pq = \tau \widetilde{p}q$$
\normalsize
\end{itemize}
\end{itemize}
\item Yet... A debated question \vspace{0.1cm}
\item Would not trade costs rather exhibit an additive structure ?  \vspace{0.1cm} \vspace{0.1cm}
\begin{itemize}
\item[-] Why would it be more costly to transport from Milan to Paris, a pair of Italian shoes at price \euro 300 than a pair of Italian shoes at price \euro 50? \vspace{0.1cm}
\item[-] Pricing shipping often includes an additive component \vspace{0.1cm}
\begin{itemize}
\item[$\star$] UPS: A \$125 fee charged for a 2 pound package from Oslo to NYC (Irrarazabal et al., 2014) \vspace{0.1cm}
\end{itemize}
\item[-] Some trade policy instruments as well (quota license price)
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\begin{itemize}
\item The structure (additive vs iceberg) of transport costs, not anecdotal \vspace{0.2cm}
\begin{itemize}
\item[-] Additive trade costs play an important role in shaping the pattern of trade flows (Alchian \& Allen, 1964) \vspace{0.1cm}
\item[-] Strong normative implications, notably w.r.t. the welfare gains of trade liberalization (Sorensen, 2014)\vspace{0.1cm}
\item[-] The additive structure of trade costs is supported by recent empirical evidence (Irrarazabal et al. 2014, Hummels \& Skiba, 2004)  \vspace{0.1cm}
\end{itemize}
\item[$\Rightarrow$] Trade costs are likely to display an additive component, but precisely... by how much? \vspace{0.2cm}
\end{itemize}
\textbf{One objective of the paper: Provide an answer to this question}

@@

\begin{frame}
\frametitle{Our paper in 3 questions (and 3 answers)}
An empirical decomposition of the structure of transport costs \vspace{0.2cm}
\begin{itemize}
\item[(1)]\textbf{ What is the size of the iceberg and the additive costs?} \vspace{0.1cm}
\item[$\Rightarrow$] Provide a quantitative measure of both, using US imports data \vspace{0.1cm}
\begin{itemize}
%\footnotesize{
\item[-] Iceberg cost: 2.5\% of the export price in air, and 3.2\% in vessel (mean value over 1974-2013) \vspace{0.1cm}
\item[-] Additive cost: 1.8\% and 2.9\% of the export price, in air / vessel  \vspace{0.2cm}
%}
\end{itemize}
\item[(2)] \textbf{ What do we lose by skipping the additive part of transport costs?} \vspace{0.1cm}
\item[$\Rightarrow$] We lose much: With the additive term included, \vspace{0.1cm}
\begin{itemize}
%\footnotesize{
\item[-] The estimated iceberg component is reduced by a factor of 2 \vspace{0.1cm}
\item[-] A significantly better ``goodness-of-fit'' \vspace{0.1cm}
%}
\end{itemize}
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{More on our database}
\begin{itemize}
\item Implications (and limitations) \vspace{0.1cm}
\begin{itemize}
\item[-] Only cover international transport costs \vspace{0.1cm}
\item[-] Among transport costs, quantitative freight costs (not those related to the time value of goods) \vspace{0.1cm}
%\item[$\Leftrightarrow$] Among the 21\% markup from AVW, $\simeq$ 11\% for freight costs (the remaining 9\% for time costs). But ``rough'' estimate. Precisely?  \vspace{0.1cm}
\end{itemize}
\item A rich database to exploit \vspace{0.1cm}
\begin{itemize}
\item[-] US imports, large time period: Broad view of international trade flows \vspace{0.1cm}
\item[-] A reliable database, already used by Hummels (2007), but on a larger period of time \vspace{0.1cm}
\item[-] Have both the import and the export prices: Estimate the levels of both the ad-valorem and the additive trade costs ($\neq$ Irrarazabal et al., 2015)
\end{itemize}
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{Time trends in transport costs \textit{per se}}
\begin{itemize}
\item Total transport costs (composition effects excluded) over time
\end{itemize}
\begin{figure}[htbp]
%\caption{Decomposing Transport costs (Yearly mean value, 3 digits)}
%\label{fig:decomp_TC_3d}
\begin{center}
\begin{tabular}{cc}
{\scriptsize (a) Iceberg alone } & {\scriptsize  (b) With additive }\\
\includegraphics[width=5cm, height=5cm]{Fig3a_TC_overtime_comp_effects_excl.pdf}
& \includegraphics[width=5cm,height=5cm]{TC_addplusmult_compeffects_excl.pdf} \\
\end{tabular}
\end{center}
\end{figure}

>>>>

\begin{frame}
\frametitle{Motivation}
\begin{itemize}
\item Trade costs have a central role in international economic analysis \vspace{0.1cm}
\begin{itemize}
\item[-] Declining over the second half of the 20$^{th}$ century (Jacks et al., 2008, Novy, 2013) \vspace{0.1cm}
\item[-] But still significant: Average international trade costs = a 74\% markup over production costs (Anderson \& Van Wincoop, 2004)
\end{itemize}
\item What exactly are ``trade costs''?  \vspace{0.1cm}
\begin{itemize}
\item[-] Transaction costs, policy costs, time costs, and transport costs \textit{per se}
\end{itemize}
\item Transport costs are a sizeable share of international trade costs \vspace{0.1cm}
\begin{itemize}
\item[-]  Amount to $\simeq 30\%$ of international trade costs  $\Leftrightarrow$ A 21\% markup over production costs  \vspace{0.1cm}
%\item[-] Elasticity of trade wrt freight costs = -3 (Behar \& Venables, 2011) \vspace{0.1cm}
\end{itemize}
\item[$\Rightarrow$] If much trade policy barriers have been removed, the transport cost component of trade costs remains sizeable \vspace{0.1cm}
\end{itemize}
\textbf{The paper: On international transport costs}

@@

\begin{frame}
\frametitle{Motivation (cont')}
\begin{itemize}
\item Standard modeling of trade costs: As an ad-valorem tax-equivalent \vspace{0.1cm}
\begin{itemize}
\item[-] As a constant percentage of the producer price per unit traded \vspace{0.1cm}
\item[$\Leftrightarrow$] The ``iceberg cost'' hypothesis (Samuelson, 1954) \vspace{0.1cm}
\begin{itemize}
\item[-] With $p$ the import price, $\widetilde{p}$ the export price, $q$ the quantity traded and $\tau >1$ the trade cost \vspace{0.1cm}
\footnotesize
$$pq = \tau \widetilde{p}q$$
\normalsize
\end{itemize}
\end{itemize}
\item Yet... A debated question \vspace{0.1cm}
\item Would not trade costs rather exhibit an additive structure ?  \vspace{0.1cm} \vspace{0.1cm}
\begin{itemize}
\item[-] Why would it be more costly to transport from Milan to Paris, a pair of Italian shoes at price \euro 300 than a pair of Italian shoes at price \euro 50? \vspace{0.1cm}
\item[-] Pricing shipping often includes an additive component \vspace{0.1cm}
\begin{itemize}
\item[$\star$] UPS: A \$125 fee charged for a 2 pound package from Oslo to NYC (Irrarazabal et al., 2014) \vspace{0.1cm}
\end{itemize}
\item[-] Some trade policy instruments as well (quota license price)
\footnotesize
$$pq = (t+\widetilde{p})q  $$
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\begin{itemize}
\item The structure (additive vs iceberg) of transport costs is not anecdotal \vspace{0.2cm}
\begin{itemize}
\item[-] Many models rely on the tractability of iceberg costs \vspace{0.1cm}
\item[-] Additive trade costs play an important role in shaping the pattern of trade flows (Alchian \& Allen, 1964) \vspace{0.1cm}
\item[-] Strong normative implications, notably w.r.t. the welfare gains of trade liberalization because they distort relative prices (Sorensen, 2014)\vspace{0.1cm}
\item[-] The additive structure of trade costs is supported by recent empirical evidence (Irrarazabal et al. 2014, Hummels \& Skiba, 2004)  \vspace{0.1cm}
\end{itemize}
\item[$\Rightarrow$] Trade costs are likely to display an additive component, but precisely... by how much? \vspace{0.2cm}
\end{itemize}
\textbf{One objective of the paper is to provide an answer to this question}

@@

\begin{frame}
\frametitle{Our paper in 3 questions (and 3 answers)}
An empirical decomposition of the structure of transport costs \vspace{0.2cm}
\begin{itemize}
\item[(1)]\textbf{ What is the size of the iceberg and the additive costs?} \vspace{0.1cm}
\item[$\Rightarrow$] Provide a quantitative measure of both, using US imports data \vspace{0.1cm}
\begin{itemize}
%\footnotesize{
\item[-] Iceberg cost: 2.5\% of the export price in air, and 3.2\% in vessel (mean value over 1974-2013) \vspace{0.1cm}
\item[-] Additive cost: 1.8\% and 2.9\% of the export price, in air / vessel  \vspace{0.2cm}
%}
\end{itemize}
\item[(2)] \textbf{ What do we lose by skipping the additive part of transport costs?} \vspace{0.1cm}
\item[$\Rightarrow$] We lose much: With the additive term included, \vspace{0.1cm}
\begin{itemize}
%\footnotesize{
\item[-] The estimated iceberg component is reduced by a factor linked to the importance of additive costs (which is kind of obvious) \vspace{0.1cm}
\item[-] A significantly better ``goodness-of-fit'' \vspace{0.1cm}
%}
\end{itemize}
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{More on our database}
\begin{itemize}
\item Implications (and limitations) \vspace{0.1cm}
\begin{itemize}
\item[-] Only cover international transport costs \vspace{0.1cm}
\item[-] Among transport costs, insurance + handling + quantitative freight costs (not those related to the time value of goods) \vspace{0.1cm}
%\item[$\Leftrightarrow$] Among the 21\% markup from AVW, $\simeq$ 11\% for freight costs (the remaining 9\% for time costs). But ``rough'' estimate. Precisely?  \vspace{0.1cm}
\end{itemize}
\item A rich database to exploit \vspace{0.1cm}
\begin{itemize}
\item[-] US imports, large time period: Broad view of international trade flows \vspace{0.1cm}
\item[-] A reliable database, already used by Hummels (2007), but on a larger period of time \vspace{0.1cm}
\item[-] Have both the import and the export prices: Estimate the levels of both the ad-valorem and the additive trade costs ($\neq$ Irrarazabal et al., 2015)
\end{itemize}
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{Time trends in transport costs \textit{per se}}
\begin{itemize}
\item Total transport costs (composition effects excluded) over time
\end{itemize}
\begin{figure}[htbp]
%\caption{Decomposing Transport costs (Yearly mean value, 3 digits)}
%\label{fig:decomp_TC_3d}
\begin{center}
\begin{tabular}{cc}
\\
 \includegraphics[width=10cm,height=7.5cm]
{graph_composition_all-eps-converted-to.pdf} \\
\end{tabular}
\end{center}
\end{figure}
